---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
editor_options: 
  chunk_output_type: console
---
```{r echo=TRUE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
```

***
## Loading and preprocessing the data
```{r}
# unzip if not yet done
if (!file.exists("./activity.csv")) {
    unzip(zipfile = "./activity.zip",
          exdir = ".")
}

dt <- data.table::fread(file = "./activity.csv")
head(dt)
```

### Pairs
```{r}
# preview
pairs(dt)
```

***
## What is mean total number of steps taken per day?
```{r}
# daily aggregation
aggregated_by_day <- dt[, .(daily_total = sum(steps, na.rm = TRUE)),
                        keyby = .(date)]

mean_by_day <- round(mean(aggregated_by_day$daily_total), 0)
```
Mean total number of steps taken per day: `r mean_by_day`

***
### Histogram I.
```{r}
# plot histogram
hist(aggregated_by_day$daily_total,
     xlab = "Number of Steps",
     main = "Daily Total Steps Oct & Nov 2012")

# calculate & plot mean
daily_mean <- round(mean(aggregated_by_day$daily_total, 
                         na.rm = TRUE), 0)
abline(v = daily_mean, col = "blue", lwd = 3)

# calculate & plot median
daily_median <- median(aggregated_by_day$daily_total,
                       na.rm = TRUE)
abline(v = daily_median, col = "red", lwd = 3)

# add text 'mean'
text(19000,
     25,
     "mean",
     col = "blue",
     cex = 1.2,
     pos = 4)

# add text 'median'
text(19000,
     20,
     "median",
     col = "red",
     cex = 1.2,
     pos = 4)
```
  
The mean of the daily number of steps taken: `r daily_mean`  
The median of the daily number of steps taken: `r daily_median`

***
## What is the average daily activity pattern?
Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
# aggregate by intervals
aggregated_by_intervals <- dt[, .(interval_mean = mean(steps, na.rm = TRUE)), 
                              keyby = .(interval)]
# plot
plot(x = aggregated_by_intervals$interval,
     y = aggregated_by_intervals$interval_mean,
     type = "l")

# determine interval with maximum number of steps
max_interval <- aggregated_by_intervals[
    order(interval_mean, 
          decreasing = TRUE), interval][1]

```
Interval `r max_interval` contains the maximum number of steps.

***
## Imputing missing values
### Nr of Missing Values
```{r}
countNAs <- sum(!complete.cases(dt))
```
The total number of missing values in the dataset (i.e. the total number of rows with NAs) is `r countNAs`.  

### Impute Missing Values & Create New Dataset
```{r}
# use the impute.knn function ======
# create a matrix based on dt
mx <- copy(dt)
# convert date into numeric value (for matrix)
mx[, date := as.numeric(date)]
mx <- as.matrix(mx)
# impute
library(impute)
mx <- impute.knn(mx)$data
# paste back imputed steps data into new working file
dt2 <- as.data.table(mx)
dt2[, date := dt$date]
dt2
```
I filled in missing values using the impute.knn() function and created new dataset: dt2.

### Histogram II.
Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r}
# daily aggregation
aggregated_by_day2 <- dt2[, .(daily_total = sum(steps, na.rm = TRUE)),
                        keyby = .(date)]
# plot histogram
hist(aggregated_by_day2$daily_total,
     xlab = "Number of Steps",
     main = "Imputed Daily Total Steps Oct & Nov 2012")

# calculate & plot mean
daily_mean2 <- round(mean(aggregated_by_day2$daily_total, 
                         na.rm = TRUE), 0)
abline(v = daily_mean2, col = "blue", lwd = 3)

# calculate & plot median
daily_median2 <- median(aggregated_by_day2$daily_total,
                       na.rm = TRUE)

abline(v = daily_median2, col = "red", lwd = 3)

# add text 'mean'
text(19000,
     25,
     "mean",
     col = "blue",
     cex = 1.2,
     pos = 4)

# add text 'median'
text(19000,
     20,
     "median",
     col = "red",
     cex = 1.2,
     pos = 4)
```
  
The total number of steps by day increased from `r daily_mean` to `r as.character(daily_mean2)`.  
The median of steps by day remained unchanged:  
Median I.: `r daily_median`  
Median II.: `r as.character(daily_median2)`

***
## Are there differences in activity patterns between weekdays and weekends?
```{r}
dt2[, weekday := weekdays(date)]
dt2[, day_type := ifelse(weekday %in% c("Saturday", "Sunday"),
                         "weekend",
                         "weekday")]
# aggregate by intervals
aggregated_by_intervals2 <- dt2[, .(steps_mean = mean(steps, na.rm = TRUE)), 
                              keyby = .(interval, day_type)]
# ggplot
library(ggplot2)
ggplot(data = aggregated_by_intervals2, 
       mapping = aes(x = interval, y = steps_mean, colour = day_type)) +
    geom_line() +
    geom_smooth()

```
  
On weekends there is less activity in the early morning hours, but more activity during the day.  

### The same again with Base R plot (just for fun)
```{r}
# weekday =====================================================================
# create vector x
vxwdi <- aggregated_by_intervals2[day_type == "weekday", interval]
# create vector y
vywdy <- aggregated_by_intervals2[day_type == "weekday", steps_mean]

# create smooth line
lo <- loess(vywdy~vxwdi)

# plot
plot(x = vxwdi,
     y = vywdy,
     type = "l",
     xlab = "time",
     ylab = "mean steps",
     main = "Mean Steps Weekday vs Weekend")

# add smoother
lines(x = vxwdi,
      y = predict(lo), 
      col='black', 
      lwd=2)

# weekend =====================================================================

# create vector x
vxwei <- aggregated_by_intervals2[day_type == "weekend", interval]

# create vector y
vywey <- aggregated_by_intervals2[day_type == "weekend", steps_mean]

# create smooth line
lowe <- loess(vywey~vxwei)

# plot
lines(x = vxwei,
     y = vywey,
     col = "red")

# add smoother
lines(x = vxwei,
      y = predict(lowe), 
      col='red', 
      lwd=2)

# add legend
legend(x = "topleft",
       legend = c("Weekday", "Weekend"),
       col = c("black", "red"),
       lty = 1,
       cex = 0.8)
```

Again, on weekends there is less activity in the early morning hours, but more activity during the day.  